Covariate analysis of UFL-Mayo-ISB RNASeq temporal cortex

### 2016-01-27 code/Rmd/Mayo_TemporalCortex_RNASeq.Rmd
Added differential expression to Mayo temporal cortex data

### Data preprocessing
```{r preprocess.data,cache=TRUE, echo=TRUE}
# Fix metadata
METADATA$AgeAtDeath = gsub("_or_above", "", METADATA$AgeAtDeath)
METADATA <- METADATA %>%
  dplyr::filter(!(ID %in% c("1923_TCX")))
```
Following five samples were considered outliers and removed 1923\_TCX
```{r preprocess.data1,cache=TRUE}
indToRetain = intersect(METADATA$ID, colnames(COUNT))

@@

# Initial design matrix with covariates under consideration
DM = getDesignMatrix(COVARIATES, Intercept = F)
```
Clustering of initial normalised data (with NULL design)
```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(VOOM.GENE_EXPRESSION$E)

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=factor(Diagnosis), shape=as.factor(Source), size=as.numeric(AgeAtDeath)))
p <- p + theme_bw() + theme(legend.position="top") 
p <- p + geom_text(aes(label= ID), size=4, hjust=0)
p

@@

# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                              DESIGN,
                                              'residual matrix of all adjusted design(voom-normalized)',
                                              isKeyPlot=TRUE)
```
```{r residualSigCovars.ALL, fig.width=25, fig.height=12}
residualSigCovars[["PC_res"]][[2]]$plotData
```
Adding "Diagnosis" back to the residuals
```{r addVarsBack}
varsToAddBackIn <- c("DiagnosisAD", "DiagnosisControl", "DiagnosisPSP", "DiagnosisPathologic Aging")
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
  t(data.matrix(DM1$design[,varsToAddBackIn]) %*% t(VOOM.ADJUSTED.FIT$coefficients[,varsToAddBackIn]))
RESIDUAL.GENE_EXPRESSION = rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, 'ensembl_gene_id') 
```

### Clustering residual gene expression
```{r decompse.normalise.data1, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(RESIDUAL.GENE_EXPRESSION[,-(1)], scale.=T)

@@

# Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
                       filters = "ensembl_gene_id", values = RESIDUAL.GENE_EXPRESSION$ensembl_gene_id,
                       mart = mart)
RESIDUAL.GENE_EXPRESSION <- left_join(RESIDUAL.GENE_EXPRESSION, Ensemble2HGNC) %>%
  dplyr::select(ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
```
### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Code
CODE <- Folder(name = 'UFL_Mayo_ISB Temporal Cortex RNASeq', parentId = parentId)
CODE <- synStore(CODE)

@@

# Store covariates
write.table(RESIDUAL.GENE_EXPRESSION, file = 'expression.tsv', sep = '\t', row.names=F, quote=F)
EXP_OBJ = File('expression.tsv', name = 'Voom Adjusted Residual Expression (Diagnosis added)', parentId = CODE$properties$id)
EXP_OBJ = synStore(EXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)
```
|  *Results*            |  *SynapseID*                    |
|  -------              |   ---------                     |
|  Covariates           |  `r COV_OBJ$properties$id`      |
|  Residual Expression  |  `r EXP_OBJ$properties$id`      |

>>>>

### Data preprocessing
```{r preprocess.data,cache=TRUE, echo=TRUE}
# Fix metadata
METADATA$AgeAtDeath = gsub("_or_above", "", METADATA$AgeAtDeath)
METADATA$Diagnosis = gsub("Pathologic Aging","Pathologic_Aging", METADATA$Diagnosis)
METADATA <- METADATA %>%
  dplyr::filter(!(ID %in% c("1923_TCX")))
```
Following five samples were considered outliers and removed 1923\_TCX
```{r preprocess.data1,cache=TRUE}
indToRetain = intersect(METADATA$ID, colnames(COUNT))

@@

# Initial design matrix with covariates under consideration
DM = getDesignMatrix(COVARIATES, Intercept = F)
```
Clustering of initial normalised data (with NULL design)
```{r decompse.normalise.data, fig.height=9, fig.width=16, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(VOOM.GENE_EXPRESSION$E)

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=factor(Diagnosis), shape=as.factor(Source), size=as.numeric(AgeAtDeath)))
p <- p + theme_bw() + theme(legend.position="top") 
# p <- p + geom_text(aes(label= ID), size=4, hjust=0)
p

@@

# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                              DESIGN,
                                              'residual matrix of all adjusted design(voom-normalized)',
                                              isKeyPlot=TRUE)
```
```{r residualSigCovars.ALL, fig.width=25, fig.height=12}
residualSigCovars[["PC_res"]][[2]]$plotData
```
Adding "Diagnosis" back to the residuals
```{r addVarsBack}
varsToAddBackIn <- c("DiagnosisAD", "DiagnosisControl", "DiagnosisPSP", "DiagnosisPathologic_Aging")
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
  t(data.matrix(DM1$design[,varsToAddBackIn]) %*% t(VOOM.ADJUSTED.FIT$coefficients[,varsToAddBackIn]))
RESIDUAL.GENE_EXPRESSION = rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, 'ensembl_gene_id') 
```

### Clustering residual gene expression
```{r decompse.normalise.data1, fig.height=10, fig.width=16, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(RESIDUAL.GENE_EXPRESSION[,-(1)], scale.=T)

@@

# Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
                       filters = "ensembl_gene_id", values = RESIDUAL.GENE_EXPRESSION$ensembl_gene_id,
                       mart = mart)
RESIDUAL.GENE_EXPRESSION <- left_join(RESIDUAL.GENE_EXPRESSION, Ensemble2HGNC) %>%
  dplyr::select(ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
```
### Differential expression analysis
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp}
# Fit contrast
contrast = makeContrasts(contrasts=c("DiagnosisControl-DiagnosisAD",
                                     "DiagnosisControl-DiagnosisPSP",
                                     "DiagnosisControl-DiagnosisPathologic_Aging",
                                     "DiagnosisPathologic_Aging-DiagnosisPSP",
                                     "DiagnosisPathologic_Aging-DiagnosisAD",
                                     "DiagnosisPSP-DiagnosisAD"), 
                         levels = colnames(VOOM.ADJUSTED.FIT$coefficients))
FIT.CONTR = contrasts.fit(VOOM.ADJUSTED.FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

# Get differnetial expression
DE = list()
p = list()
for (i in 1:6){
  DE[[i]] = topTable(FIT.CONTR, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1])
  DE[[i]] = rownameToFirstColumn(DE[[i]], 'ensembl_gene_id') %>%
    left_join(Ensemble2HGNC)
  p[[i]] <- ggplot(DE[[i]], aes(x = logFC, y = -log10(adj.P.Val))) + geom_point()
  p[[i]] <- p[[i]] + geom_hline(yintercept = -log10(0.05), color = 'red') 
  p[[i]] <- p[[i]] + ggtitle(gsub('Diagnosis','',colnames(contrast)[i]))
}
multiplot(plotlist=p, cols = 3)
names(DE) = gsub('Diagnosis','',colnames(contrast))

DE = ldply(DE, .id = 'Comparison')
tmp = DE %>% 
  dplyr::filter(adj.P.Val <= 0.05) %>%
  group_by(Comparison) %>%
  dplyr::summarize(nDiffExpGenes = length(unique(hgnc_symbol)),
                   DiffExpGenes = paste(unique(hgnc_symbol), collapse = ','))
kable(tmp)
```

### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Code
CODE <- Folder(name = 'UFL_Mayo_ISB Temporal Cortex RNASeq', parentId = parentId)
CODE <- synStore(CODE)

@@

# Store covariates
write.table(RESIDUAL.GENE_EXPRESSION, file = 'expression.tsv', sep = '\t', row.names=F, quote=F)
EXP_OBJ = File('expression.tsv', name = 'Voom Adjusted Residual Expression (Diagnosis added)', parentId = CODE$properties$id)
EXP_OBJ = synStore(EXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)

# Store differential expression results
write.table(DE, file = 'diffExpression.tsv', sep = '\t', row.names=F, quote=F)
DEXP_OBJ = File('diffExpression.tsv', name = 'Differential Expression Analysis', parentId = CODE$properties$id)
DEXP_OBJ = synStore(DEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)
```
|  *Results*            |  *SynapseID*                    |
|  -------              |   ---------                     |
|  Covariates           |  `r COV_OBJ$properties$id`      |
|  Residual Expression  |  `r EXP_OBJ$properties$id`      |
|  Differential Expression  |  `r DEXP_OBJ$properties$id`      |
